---
title: "Not coordinated departures/returns in relation with foraging areas"
author: "Camila Calderon"
date: "2024-07-25"
output: html_document
---

### Load libraries

```{r load packages, warning=FALSE, message=FALSE}
pacman::p_load(dplyr, lubridate, ggplot2, data.table, plyr, raster, ggpubr, emmeans, tidyverse, factoextra)
```

### Load data

```{r}
load(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/PlosOne/data/females_HMMbehaviors.RData")
```

### Extract foraging points from data set of la Gruta colony in February

Here I am adding a variable that counts the length of consecutive commuting and foraging points. This will be used to filter commuting points that have 5 or more points in a row.

```{r, results=FALSE}
hast_df <- females_bocas
hast_df$n <- NA
r <- rle(hast_df$behav)

for(i in 1:length(r$lengths)){
  #if(r$values[i] == "commuting"){
      start <- {}
      end <- {}
      try(start <- sum(r$lengths[1:(i-1)])+1)
      if(i == 1) start <- 1
      end <- sum(r$lengths[1:i])

      hast_df$n[start:end] <- r$lengths[i]
}

# data frame with only tracking from February
bats_feb <- hast_df %>%
 filter(date > "2022-01-01" & date < "2022-03-01")

# split data frame by ID
bats_feb_split <- split(bats_feb, bats_feb$ID)

```

### Select commuting paths to extract departure times

```{r}
# filter only commuting with 5 or more points in a row
bats_commufeb <- bats_feb  %>%  
  filter(behav=="commuting" & n>=5)

# take the first row of each id date in the commuting tracks
commu_feb_dep <- bats_commufeb %>%
  group_by(ID_batday) %>%
  slice(which.min(1:n()))

# remove 2D507235_G_2022−01−31
commu_feb_dep <- commu_feb_dep[commu_feb_dep$ID!="2D507235_G_2022-01-31",]

# split data frame by date
commu_feb_dep_date <- split(commu_feb_dep, f=commu_feb_dep$date)

```

### Extract return times

```{r}
# calculating return foraging locations
commu_feb_ret <- bats_commufeb %>%
  group_by(ID) %>%
  slice(which.max(1:n()))

# remove 2D507235_G_2022−01−31
commu_feb_ret <- commu_feb_ret[commu_feb_ret$ID!="2D507235_G_2022-01-31",]

# split by date
commu_feb_ret_date <- split(commu_feb_ret, commu_feb_ret$date)
```

### Calculate pairwise time difference in departures by date, between and within groups

Here we include days with more than 2 individuals

```{r}
pair.depdiff <-  rbindlist(lapply(commu_feb_dep_date[c(1:11)], FUN=function(t){
  combi <- t(combn(t$tag_local_identifier, 2)) # all unique pairwise combinations of individuals
  date <- unique(date(t$timestamps))#UTC time
  combi <- data.frame(bat1=combi[,1], bat2=combi[,2], date=date, stringsAsFactors=F) # set up a new data frame
  combi <- combi[combi$bat1!=combi$bat2,]# just in case that a single bat has two readings for the same timestamp
  
  combi$Deptimediff <- unlist(lapply(1:nrow(combi), function(i){ # calculate time difference between all pairs (in seconds)
    pos.bat1 <- t[t$tag_local_identifier==combi$bat1[i], c('timestamp')] # time for first bat
    pos.bat2 <- t[t$tag_local_identifier==combi$bat2[i], 'timestamp'] # time for second bat
    return(difftime(pos.bat1$timestamp, pos.bat2$timestamp, units="mins"))})) # time in minutes
  return(combi)
  }))


# create a new table for time differences 
pairwisetime_table <- pair.depdiff %>%
  mutate(bats= paste(bat1, bat2, sep="_")) %>%
  mutate(Deptimediff= abs(pair.depdiff$Deptimediff), group1= str_sub(bat1, 10,10), group2= str_sub(bat2, 10,10)) %>%
  mutate(samegroup=ifelse(group1 == group2, "yes", "no"))

```


### Calculate pairwise time difference in returns by date, between and within groups

```{r}
pair.retdiff <-  rbindlist(lapply(commu_feb_ret_date[c(1:11)], FUN=function(t){
  combi <- t(combn(t$tag_local_identifier, 2)) # all unique pairwise combinations of individuals
  date <- unique(date(t$timestamps))#UTC time
  combi <- data.frame(bat1=combi[,1], bat2=combi[,2], date=date, stringsAsFactors=F) # set up a new data frame
  combi <- combi[combi$bat1!=combi$bat2,]# just in case that a single bat has two readings for the same timestamp
  
  combi$Deptimediff <- unlist(lapply(1:nrow(combi), function(i){ # calculate time difference between all pairs (in seconds)
    pos.bat1 <- t[t$tag_local_identifier==combi$bat1[i], c('timestamp')] # time for first bat
    pos.bat2 <- t[t$tag_local_identifier==combi$bat2[i], 'timestamp'] # time for second bat
    return(difftime(pos.bat1$timestamp, pos.bat2$timestamp, units="mins"))})) # time in seconds
  return(combi)
  }))

# insert return time difference to the pairwisetime_table
pairwisetime_table$rettimediff <-abs(pair.retdiff$Deptimediff)
pairwisetime_table$groups=paste(pairwisetime_table$group1, pairwisetime_table$group2,sep="")

# change the name of the groups to have only the combination GD
pairwisetime_table$groups[which(pairwisetime_table$groups=="GD")] <- "DG"
```

### Test pairwise difference in departures and returns between and within individuals

Here I created a threshold of foraging together or not based on a 2 minute time difference, according to the GPS fix resolution

```{r}
# define the threshold for departures
threshold <- 2

# create the new column using ifelse()
pairwisetime_table$together <- ifelse(pairwisetime_table$Deptimediff >= threshold, 0, 1)
pairwisetime_table$together2 <- ifelse(pairwisetime_table$rettimediff >= threshold, 0, 1)

# how many bats forage together based on this threshold?
table(pairwisetime_table$together)
table(pairwisetime_table$together2)

# how many days have less than 3 points?
table(pairwisetime_table$date)
```


### Test differences in departure and returns by groups
```{r}
library(lme4)
# making sure there are no zeros in the data frame
pairwisetime_table$Deptimediff[which(pairwisetime_table$Deptimediff==0)] <- 0.001

# run a gamma model for the time difference between groups
m_dep <- glmer(Deptimediff~groups + (1|bats)+(1|date), data=pairwisetime_table, family = Gamma("log"))

summary(m_dep)
anova(m_dep)
# there is no departure time difference between or within groups 

# run a gamma model for the time difference between groups
m_ret <- glmer(rettimediff~groups + (1|bats)+(1|date), data=pairwisetime_table, family = Gamma("log"))

summary(m_ret)
anova(m_ret)
```

### Plot departures and return with predicted values 
```{r}
#set same group variable as factor
pairwisetime_table$samegroup_f <- factor(pairwisetime_table$samegroup, levels=c("yes","no"))

# predicted values
dep_r <- predict(m_dep, type = "response")

# combine predicted values with original data
pairwisetime_table$dep_r <- dep_r

dep <- ggplot(data=as.data.frame(pairwisetime_table), aes(x=samegroup_f, y=Deptimediff, color=groups, alpha=0.8))+
  #geom_boxplot()+
  #scale_x_discrete(breaks=c("DD", "DG", "GG"), labels=c("F1", "F1F2", "F2"))+
  geom_jitter(aes(y = Deptimediff), position = position_jitterdodge(jitter.width = 0.3, jitter.height = 0.05))+
  scale_colour_manual(aes(y= dep_r),values = c("#440154FF","grey", "#FDE725FF")) +
  stat_summary(aes(y= dep_r,group=groups),fun = mean, color = "black",
    position = position_dodge(.8)) +
  stat_summary(aes(group=groups),
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "black",
    width = .3,
    position = position_dodge(.8)
  )+
  ylab("departure time \n difference (mins)")+
  xlab("same group")+
  theme_classic()+
  guides(color="none", alpha="none")


# predicted values
ret_r <- predict(m_ret, type = "response")

# combine predicted values with original data
pairwisetime_table$ret_r <- ret_r

# plot return time difference
ret <- ggplot(data=as.data.frame(pairwisetime_table), aes(x=samegroup_f, y=rettimediff, color=groups, alpha=0.8))+
  #geom_boxplot()+
  #scale_x_discrete(breaks=c("DD", "DG", "GG"), labels=c("F1", "F1F2", "F2"))+
  geom_jitter(aes(y = rettimediff), position = position_jitterdodge(jitter.width = 0.3, jitter.height = 0.05))+
  scale_colour_manual(aes(y=ret_r), values = c("#440154FF","grey", "#FDE725FF")) +
  stat_summary(aes(y=ret_r, group=groups),fun = mean, color = "black",,
    position = position_dodge(.8)) +
  stat_summary(aes(group=groups),
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color = "black",
    width = .3,
    position = position_dodge(.8)
  )+
  ylab("return time \n difference (mins)")+
  xlab("same group")+
  theme_classic()+
  guides(color="none", alpha="none")

ret
```


### Plot final figure of departure, return and foraging area coordination

```{r}
library(patchwork)


Fig3bc <- (((dep|ret) + plot_layout(axis_titles = "collect", guides="collect")) + plot_annotation(tag_levels = "A") & 
               theme(legend.position='bottom', plot.tag  = element_text(face = 'bold', size=24), axis.title = element_text(face = 'bold', size=24), axis.title.x = element_text(size = 24), axis.title.y=element_text(size=24), axis.text.y = element_text(size=14), axis.text.x = element_text(size=14)))#pairdistplot

Fig3bc 
```

## Check pairwise time difference for female bats in 2016

Here we filter commuting points for females bats in la Gruta in 2016, extract their departures and returns to the roost.

```{r, results=FALSE}
# filter data of 2016
bats_2016 <- hast_df %>% 
 filter(date(timestamp) < "2021-01-01" )

# remove individduals with some incomplete data
bats_2016 <- bats_2016 %>%
  filter(tag_local_identifier!="74DDA80" & tag_local_identifier!="2016030705")

# filter only commuting with more than 5 points in a row
bats_commu2016 <- bats_2016  %>% 
  filter(behav=="commuting" & n>=5)

# plot commuting
ggplot(aes(x=location_long, y=location_lat, col=tag_local_identifier), data=bats_commu2016) + geom_point()

# take the first row of each id date in the commuting tracks
commu_2016_dep <- bats_commu2016 %>%
  group_by(ID) %>%
  slice(which.min(1:n()))

# split data frame by date
commu_2016_dep_split <- split(commu_2016_dep, f=commu_2016_dep$date)
```

### Calculate pairwise time difference in departures in 2016 
```{r }
# calculate pairwise time difference in departures between individuals
pair.depdiff_2016 <-  rbindlist(lapply(commu_2016_dep_split, FUN=function(t){
  combi <- t(combn(t$tag_local_identifier, 2)) # all unique pairwise combinations of individuals
  date <- unique(date(t$timestamp))
  combi <- data.frame(bat1=combi[,1], bat2=combi[,2], data=date, stringsAsFactors=F) # set up a new data frame
  combi <- combi[combi$bat1!=combi$bat2,]# just in case that a single bat has two readings for the same timestamp
  
  combi$Deptimediff <- unlist(lapply(1:nrow(combi), function(i){ # c
    pos.bat1 <- t[t$tag_local_identifier==combi$bat1[i], c('timestamp')] # time for first bat
    pos.bat2 <- t[t$tag_local_identifier==combi$bat2[i], 'timestamp'] # time for second bat
    return(difftime(pos.bat1$timestamp, pos.bat2$timestamp, units="mins"))})) # time difference between bats (in seconds)
  return(combi)
  }))

# add absolute values to the departure time difference
pair.depdiff_2016 <- pair.depdiff_2016 %>%
  mutate(Deptimediff= abs(pair.depdiff_2016$Deptimediff)) %>%
  mutate(cave_group= "lagruta", bats=paste(bat1, bat2, sep="."))

mean(pair.depdiff_2016$Deptimediff)
sd(pair.depdiff_2016$Deptimediff)
```


### Calculate pairwise time difference of returns in 2016

```{r}
# extract return times
commu_2016_ret <- bats_commu2016 %>%
  group_by(ID) %>%
  slice(which.max(1:n()))

# split by date
commu_2016_ret_split <- split(commu_2016_ret, commu_2016_ret$date)

# calculate pairwise time difference in returns between individuals 
pair.retdiff_2016 <-  rbindlist(lapply(commu_2016_ret_split, FUN=function(t){
  combi <- t(combn(t$tag_local_identifier, 2)) # all unique pairwise combinations of individuals
  date <- unique(date(t$timestamp))
  combi <- data.frame(bat1=combi[,1], bat2=combi[,2], data=date, stringsAsFactors=F) # set up a new data frame
  combi <- combi[combi$bat1!=combi$bat2,]# just in case that a single bat has two readings for the same timestamp
  
  combi$Deptimediff <- unlist(lapply(1:nrow(combi), function(i){ # calculate geographic distance between all pairs (in meters!!)
    pos.bat1 <- t[t$tag_local_identifier==combi$bat1[i], c('timestamp')] # time for first bat
    pos.bat2 <- t[t$tag_local_identifier==combi$bat2[i], 'timestamp'] # time for second bat
    return(difftime(pos.bat1$timestamp, pos.bat2$timestamp, units="mins"))})) # time difference between bats (in minutes)
  return(combi)
  }))

# add absolute values to the return time difference
pair.retdiff_2016 <- pair.retdiff_2016 %>%
  mutate(pair.retdiff= abs(pair.retdiff_2016$Deptimediff)) %>%
  mutate(cave_group= "lagruta",bats=paste(bat1, bat2, sep="."))

mean(pair.retdiff_2016$pair.retdiff)
sd(pair.retdiff_2016$pair.retdiff)
```


