---
title: "Foraging area/overlap between groups in 2022"
output: html_document
date: "2024-07-25"
Author: Camila Calderon
---

### Load packages

```{r setup, include=FALSE}
pacman::p_load(ctmm, tidyverse, lubridate, ggplot2, move, doBy, ggmap, spatialrisk, plotly)
```

### Load data

```{r load, results="hide", include=TRUE}
load(file="~/ownCloud/PhDLife/P.hastatus/Thesis/paper2/analysis/data/females_HMMbehaviors.RData")

unique(females_bocas$cave)

# adding counts to each consecutive point of foraging and commuting
females_bocas$n <- NA

r <- rle(females_bocas$behav)

for(i in 1:length(r$lengths)){
  #if(r$values[i] == "commuting"){
  start <- {}
  end <- {}
  try(start <- sum(r$lengths[1:(i-1)])+1)
  if(i == 1) start <- 1
  end <- sum(r$lengths[1:i])
  
  females_bocas$n[start:end] <- r$lengths[i]
}


# foraging points from la gruta dry 2016 and 2022 where they stay longer than 8 minutes
females_foraging <- females_bocas %>%
  filter((behav=="foraging" & n>=4))

# plot foraging points
mapview::mapView(females_foraging, xcol="location_long", ycol="location_lat", zcol="tag_local_identifier", legend=F, crs="EPSG:4326")

```

### Remove foraging points inside Isla Colón

```{r pressure, results="hide", message=FALSE, warning=FALSE}

# set coordinates in middle of the roost - La Gruta
roostlagruta <- data.frame(location.lat=9.396448, location.long=-82.271541)

# selecting foraging points based on the radios by taking la Gruta as a reference point
roostpointGruta <- points_in_circle(females_foraging, roostlagruta$location.long,roostlagruta$location.lat, radius = 6500, lon=location_long,lat=location_lat)

# remove island foraging points from the data frame
foragingall_noisla <- females_foraging[!(females_foraging$event_id %in% roostpointGruta$event_id),] 

# plot without in island points
plot(foragingall_noisla$location_long, foragingall_noisla$location_lat)

# check unique IDs
unique(foragingall_noisla$date)

# create a list
bats.cave_ls <- split(foragingall_noisla, foragingall_noisla$ID)

# filter to have only id_days with more than 10 points to be able to calculate AKDE
bats.cave_ls <- Filter(function(bats.cave_ls) nrow(bats.cave_ls) > 10, bats.cave_ls)

# convert to data frame again
bats.cave_clean <- do.call(rbind, bats.cave_ls)

# plot foraging points of filtered data frame
ggplot(aes(x=location_long, y=location_lat, color=year_cave), data=bats.cave_clean)+geom_point()

```

### Convert to telemetry object using only 2022 data

```{r convert to telemetry}
# data frame with only tracking from February
bats_feb <- bats.cave_clean %>% 
 filter(date> "2022-01-01" & date < "2022-03-01")
names(bats_feb)

# save(bats_feb_idday, file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/data/bats_feb_idday.RData")

## convert to telemetry object
# here we remove columns with different id names to use the column tag_local_identifier to convert to telemetry object
tele.feb <- as.telemetry(bats_feb[,c(-12,-24,-34,-42,-44,-46,-64)])
tele.feb[[1]]@info$projection
```

### Calculate fits for the telemery object

```{r foraging area overlaps}
# include error
uere(tele.feb) <- 20

# include two-point equidistant projection
ctmm::projection(tele.feb) <- median(tele.feb, k=2)

# fits calculation
FITS <- lapply(tele.feb, function(x){
  print(x)
  SVF <- variogram(x)#, dt = c(2, 1440 %#% "minutes"))
  GUESS <- ctmm.guess(x, CTMM=ctmm(error=TRUE), interactive=FALSE)
  FITS <- ctmm.select(x,GUESS,trace=3,cores=2, verbose=T)
  return(FITS)
})

# check for projection
projection(FITS) 

# rename fits according to the telemetry object
names(FITS) <- names(tele.feb)

# save fits
save(FITS,file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/PlosOne/data/fits_all_Feb_20240412.RData")
#load(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/PlosOne/data/fits_all_Feb_20240412.RData")
```

### Calculate UDs for all individuals

```{r}
# calculates UDS
# shared grid for UD calculations
ex <- extent(tele.feb)
ex$x <- round(ex$x/500 + c(-10,10))*500
ex$y <- round(ex$y/500 + c(-10,10))*500

# check projection
proj.all <- tele.feb[[1]]@info$projection

# set grid
grid <- list(dr=c(500,500), extent=raster:::extent(ex$x, ex$y))

# i=1

# UD calculation
UDS <- lapply(1:length(tele.feb), function(i){
  id <- names(tele.feb[i])
  akde(tele.feb[[i]],FITS[[i]], weights=TRUE, grid=grid)
})

# rename UDs according to the telemetry object
names(UDS) <- names(tele.feb)

# save UDs
save(UDS,file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/PlosOne/data/UDS_all_Feb_20240412.RData")
#load(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/PlosOne/data/UDS_all_Feb_20240412.RData")

# plot UDs
plot(UDS)  

```


### Overlap between and within groups from 2022

Here I calculate the Bhattacharyya coefficient which is a useful measure of similarity between distributions. It is roughly speaking the ratio of the intersection area to the average individual area, but it is a direct comparison between the density functions and does not require an arbitrary quantile to be specified. When applied to ctmm objects, this function returns the overlap of the two Gaussian distributions. When applied to aligned UD objects with corresponding movement models, this function returns the overlap of their (autocorrelated) kernel density estimates (from ctmm package).

I calculate the overlap by individuals pooling all days of one individual together.

```{r}
# calculate overlap
BA_ud95 <- overlap(UDS, level=.95) 
BA_ud50 <- overlap(UDS, level=.50) 
BA_ud20 <- overlap(UDS, level=.20) 

# convert to data frame
library(reshape2)
BA_ud95_df <- melt(BA_ud95$CI[,,"est"])

# remove duplicates
BA_ud95_df <- BA_ud95_df[!duplicated(BA_ud95_df$value),]

# remove comparison between same ID
BA_ud95_df <- BA_ud95_df[BA_ud95_df$Var1!=BA_ud95_df$Var2,]

# add group and id to the table
BA_ud95_df <- BA_ud95_df %>%
   mutate(id1 = str_sub(Var1, 1,10),id2 = str_sub(Var2, 1,10),
         group1 = str_sub(Var1,10,10), group2 = str_sub(Var2,10,10),
         groups=paste(group1,group2, sep="_"))

# change groups
BA_ud95_df$groups[which(BA_ud95_df$groups=="D_G")] <- "G_D"

# verify groups
unique(BA_ud95_df$groups)
```

### Plot overlap between groups

```{r}
# Fig 4B overlap error bar plot
overlap_feb <- ggplot(aes(x=groups, y=value, color=groups), data=BA_ud95_df) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2))+
  scale_x_discrete(breaks=c("D_D", "G_D","G_G"),labels=c("F1", "F1F2","F2"))+
  scale_colour_manual(values = c("#440154FF","grey", "#FDE725FF")) +
  stat_summary(fun = mean, color="black") +
  stat_summary(
    fun.min = function(x) mean(x) - sd(x), 
    fun.max = function(x) mean(x) + sd(x), 
    geom = "errorbar",
    color="black",
    width = .3
  ) +
  ylab("BA")+
  theme_classic()+
  theme(legend.position = "none")

overlap_feb

ggsave(file="~/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/PlosOne/figures/Fig4B.pdf")
```

### Test overlap between and within groups using a logistic regression

```{r}
library(lme4)
library(glmmTMB)
library(DHARMa)

head(BA_ud95_df)
str(BA_ud95_df)

# create dyad column and samegroup column
BA_ud95_df <- BA_ud95_df %>%
  mutate(samegroup=ifelse(group1 == group2, "yes", "no"), dyad=paste(Var1,Var2, sep="_"))


# removing zero values
BA_ud95_df$value[which(BA_ud95_df$value==0)] <- 0.001
BA_ud95_df$samegroup <- factor(BA_ud95_df$samegroup, levels = c("yes","no"))

# model
mod_ba_beta <- glmmTMB(value ~ samegroup + (1|dyad),
                       data = BA_ud95_df,
                       family = beta_family(link = "logit"))

# summary of the model
summary(mod_ba_beta)

# convert values of the model to percentages/proportions
logit_to_prob <- function(logit) {
  exp(logit) / (1 + exp(logit))
}

# assuming your model intercept and coefficient for samegroupno are as follows
intercept_logit <- -1.4773
coef_samegroupno <- 0.1339

# Calculate probabilities
prob_intercept <- logit_to_prob(intercept_logit)  # probability when dyads are in the same group
prob_samegroupno <- logit_to_prob(intercept_logit + coef_samegroupno)  # not in the same group

# print the results
print(paste("Probability when dyads are in the same group:", prob_intercept))
print(paste("Probability when dyads are not in the same group:", prob_samegroupno))

```
 
### Plot figure 4A

```{r plotting areas in a map}
library(sf)
######################################################################
############  Plotting and AKDE HR on a map for February ##############
######################################################################

# remove id in which AKDE was not well calculated
UDS_sub <- UDS[names(UDS) %in% c("22507B4E_D")==FALSE]

# change UD list to sf object (mean contour only) and switch to lat long coords
UD_sf <- UDS_sub  %>% 
  purrr::map(ctmm::as.sf, level.UD = 0.95) %>% 
  reduce(rbind) %>% 
  mutate(id = rep(names(UDS_sub), each = 3),
         group = str_sub(id,10)) %>% #year = as.numeric(str_sub(id,4,7)
  filter(str_detect(name, "est")) %>% 
  dplyr::select(id, group, geometry) %>% 
  st_transform(crs = "+proj=longlat +datum=WGS84")

# transform confidence intervals to an sf object
CI_sf <-  UDS_sub  %>% 
  purrr::map(ctmm::as.sf, level.UD = 0.95) %>% 
  reduce(rbind) %>% 
  mutate(id = rep(names(UDS_sub), each = 3),
         group = str_sub(id,10)) %>% #year = as.numeric(str_sub(id,4,7)
  filter(!str_detect(name, "est")) %>% 
  dplyr::select(id,group, geometry) %>% 
  st_transform(crs = "+proj=longlat +datum=WGS84")


# remove id in which AKDE was not well calculated
tele.feb_sub <- tele.feb[names(tele.feb) %in% c("22507B4E_D")==FALSE]

# first change data into sf object as well to plot with UDs
DATA_sf <- tele.feb_sub %>%
  purrr::map(ctmm::as.sf) %>% 
  reduce(rbind) %>% 
  mutate(id = identity,
         group = str_sub(id,10)) %>% 
  dplyr::select(id,group, geometry) %>% 
  st_transform(crs = "+proj=longlat +datum=WGS84")


# register stamen map
register_stadiamaps("myownkey") 

# create a bounding box for the map
e <- make_bbox(location_long, location_lat, data = bats_feb, f = .70)

# figure 4A - map by group
library(scales)
show_col(viridis_pal()(5))

Uds_map_feb <- get_stadiamap(e, zoom = 12, maptype = "stamen_terrain") %>% ggmap()+
  geom_sf(data = CI_sf, 
          aes(fill=group, color=group, alpha = 0.02, size = 0.3),
          #color = "black", 
          linetype = "dashed", 
          inherit.aes = FALSE) +
          #alpha = 0.02, 
          #size = 0.3) +
  geom_sf(data = UD_sf,
          aes(fill=group, color=group),
          inherit.aes = FALSE, 
          alpha = 0.3, 
          size = 0.3) +
  geom_sf(data = DATA_sf,
          aes(color=group, alpha= 0.03),
          size = 0.5, 
          inherit.aes = FALSE) +
  scale_colour_manual(values = c("#440154FF", "#FDE725FF")) +
  scale_fill_manual(values = c("#440154FF", "#FDE725FF")) +
  ggspatial::annotation_scale(location = "bl", 
                              width_hint = 0.3,
                              height = unit(4,'pt'),
                              style = 'ticks', text_cex =1) +
# coord_sf(xlim=c(-85.39,-85.35), ylim=c(10.495, 10.542)) + # can use this to trim the map to particular groups
  guides(color=guide_legend(override.aes=list(size = 3)))+
  xlab("longitude")+
  ylab("latitude")+
  theme(legend.position="none", 
        legend.title = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank())
  #facet_wrap(~group) # could use group instead, or remove if you would rather all in one plot
Uds_map_feb 

ggsave("/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/PlosOne/figures/Fig4A.pdf") 
```

# Plot figure 4 - composite

```{r}
library(patchwork)

((Uds_map_feb + overlap_feb) + plot_annotation(tag_levels = "A") & 
               theme(plot.tag  = element_text(face = 'bold', size=24), axis.title = element_text(face = 'bold', size=20), axis.title.x = element_text(size = 20), axis.title.y=element_text(size=20), axis.text.y = element_text(size=10), axis.text.x = element_text(size=10)))

ggsave("/Users/ccalderon/ownCloud/PhDLife/P.hastatus/Thesis/Paper2/analysis/results/figures/Fig4.pdf", width=15, height=7) 
```
